GPROF-NN: A neural-network based implementation of the Goddard Profiling Algorithm

### 2021-12-11 article/manuscript.tex
Merge.
Manuscript writing.

\title{GPROF-NN: A neural-network based implementation of the Goddard Profiling Algorithm}

\documentclass[12pt]{article}

\usepackage{amsmath}
\begin{document}
\maketitle

\section{Introduction}
\section{Methods}

This section describes the original GPROF algorithm as well as the two newly
developed neural-network-based versions.

\subsection{The original GPROF algorithm}

The original GPROF algorithm is based on a Bayesian formulation of the retrieval
problem. Given an observation $\mathbf{y}$, the result of the retrieval is
defined as the posterior distribution $P(\mathbf{x} | \mathbf{y})$ of the
atmospheric state $\mathbf{x}$.

To calculate the posterior $P(\mathbf{x} | \mathbf{y})$ the algorithm uses a
database consisting of pairs
$(\mathbf{y}_1, \mathbf{x}_1), \ldots (\mathbf{y}_N, \mathbf{x}_N)$
of observations $\mathbf{y}_i$ and corresponding states
$\mathbf{x}_i$. Assuming that the database describes the true a priori
distribution of atmospheric states $\mathbf{x}$ sufficiently well, the expected
value w.r.t to the posterior distribution $P(\mathbf{x} | \mathbf{x})$ of the
atmospheric state $\mathbf{x}$ can be approximated using
\begin{align}
  \int_{\mathbf{x} } f(\mathbf{x})\: d\mathbf{x} \approx
  \frac{\sum_i f(\mathbf{x}) P(\mathbf{y}|\mathbf{x}_i)}{\sum_i P(\mathbf{y}|\mathbf{x})}
\end{align}
The conditional probability of the observation given an atmospheric state
from the database, $P(\mathbf{y}|\mathbf{x}_i)$, is approximated using a
Gaussian error model
\begin{align}
  P(\mathbf{y}|\mathbf{x}_i) = \frac{1}{\sqrt(2\pi\text{det}(\mathbf{S}^{-1}))}
  \exp \left \{
  - \frac{1}{2}
  (\mathbf{y} - \mathbf{y}_i})^T \mathbf{S}^{-1}  (\mathbf{y} - \mathbf{y}_i)}{2})
  \right }
\end{align}

This approach can be extended to also estimate other statistics of the a
posteriori distribution such as its quantiles.

>>>>

\RequirePackage[l2tabu,orthodox]{nag} % turn on warnings because of bad style
\documentclass[a4paper,11pt,bibtotoc]{scrartcl}
\usepackage{natbib}

\usepackage[utf8]{inputenc}

\usepackage[T1]{fontenc}        % Tries to use Postscript Type 1 Fonts for better rendering
\usepackage{lmodern}            % Provides the Latin Modern Font which offers more glyphs than the default Computer Modern
\usepackage{amsmath} % Provides all mathematical commands
\usepackage{amsfonts} % Provides all mathematical commands

\usepackage{hyperref}           % Provides clickable links in the PDF-document for \ref
\usepackage{grffile}            % Allow you to include images (like graphicx). Usage: \includegraphics{path/to/file}
\usepackage{subcaption}

% Allows to set units
\usepackage[ugly]{units}        % Allows you to type units with correct spacing and font style. Usage: $\unit[100]{m}$ or $\unitfrac[100]{m}{s}$

% Additional packages
\usepackage{url}                % Lets you typeset urls. Usage: \url{http://...}
\usepackage{breakurl}           % Enables linebreaks for urls
\usepackage{xspace}             % Use \xpsace in macros to automatically insert space based on context. Usage: \newcommand{\es}{ESPResSo\xspace}
\usepackage{xcolor}             % Obviously colors. Usage: \color{red} Red text
\usepackage{booktabs}           % Nice rules for tables. Usage \begin{tabular}\toprule ... \midrule ... \bottomrule

% Source code listings
\usepackage{listings}           % Source Code Listings. Usage: \begin{lstlisting}...\end{lstlisting}
\lstloadlanguages{python}       % Default highlighting set to "python"

\renewcommand{\familydefault}{\sfdefault}
\newcommand\todo[1]{\textcolor{red}{\textbf{TODO :: }#1}}
\newcommand\note[1]{\textcolor{green}{\textbf{NOTE :: }#1}}

\begin{document}

\title{GPROF-NN: A neural-network based implementation of the Goddard Profiling Algorithm}
\date{\today}

\maketitle

\section{Introduction}

The Goddard Profiling Algorithm (GPROF, \citet{kummerow15}) is the operational
precipitation retrieval algorithm for the passive microwave (PMW) observations
from the constellation of satellites of the Global Precipitation Measurement
(GPM, \citet{hou14}) mission, whose objective is to provide global measurements
of precipitation at a temporal resolution of 3 hours. The precipitation that is
retrieved using GPROF is used directly by meteorologists and climate scientists
and serves as input for GPM level 3 retrieval products. The algorithm thus
constitutes an essential component of the system of global observations that
enables monitoring of the hydrological cycle for the benefit of science and
society.

The development of GPROF was originally motivated by the Tropical Rainfall
Measurement Mission (TRMM, \citet{simpson96}), the precursor of the GPM mission,
and thus dates back almost 30 years \citep{kummerow94_a, kummerow94_b}. Due to
the conceptual and computational complexity of simulating PMW observations of
clouds and precipitation, the algorithm was and remains based on a retrieval
database consisting of observations and corresponding profiles of hydrometeors
and precipitation rates. Nonetheless, the algorithm has undergone several
updates since its conception: Methodologically, the most fundamental
modification was the introduction of the Bayesian retrieval scheme in
\citet{kummerow96}, which is used in the algorithm until today. Following this,
algorithm updates were mostly focused on improving the retrieval database and
the incorporation of ancillary data into the retrieval. While the first version
of GPROF still used hand crafted hydrometeor profiles to generate the retrieval
database, these were soon replaced by profiles from a meso-scale weather model
\citep{kummerow96}. An important improvement was the replacement of the
model-derived database by an observationally-generated database for the GPROF
2010 algorithm \citep{kummerow11, kummerow15}, which helped reduce errors caused
by misrepresentation of atmospheric states in the database. The 2014 version of
GPROF \citep{kummerow15} introduced the first fully-parametric version of the
algorithm, which was designed to be applicable to all sensors of the GPM
constellation. This version of GPROF became the operational PMW
precipitation retrieval of the GPM mission.

This study focuses on the computational method that is used to produce the
retrieval results from the retrieval database used by GPROF. Since its
introduction in \citet{kummerow96}, the currently used Bayesian method has not
received much consideration, mainly because the database and the incorporation
of ancillary data were deemed to be more relevant for improving the accuracy of
the retrieval. However, two disadvantages of the current retrieval method have
become apparent with the introduction of the much larger,
observationally-generated retrieval databases into the algorithm
\citep{elsaesser15}: Firstly, the retrieval database must be compressed into
self-similar clusters to make calculation of the retrieval results sufficiently
fast. This lossy compression may limit the extent to which the current algorithm
can benefit from the size and representativeness of observationally generated
retrieval databases. Secondly, the current retrieval method requires artificial
inflation of the assumed database uncertainties to produce accurate results.
Since there is no principled way to calculate these uncertainties, they need to
be tuned heuristically for each sensor.

With the aim of overcoming these limitations, this study investigates the
benefits of using the GPROF retrieval database to train a neural network to
retrieve precipitation and hydrometeor profiles. Since the retrieval database
has grown to a size of several hundred million entries it is perfectly suited
for the application of deep neural networks, which scale very well to large
amounts of data and are capable of learning highly complex relationships from
them. In addition to this, a neural network based implementation has the
advantage of allowing the integration of spatial information into the retrieval,
which is not easily feasible with the current method. This study thus aims to
answer the following two questions:
\begin{enumerate}
\item Can a deep learning based retrieval method with identical inputs improve
  the accuracy of retrieved surface precipitation and vertical hydrometeor
  profiles?
  \item Can the incorporation of spatial information into the retrieval help to
    improve the retrievals?
\end{enumerate}

To answer these questions, this study presents two novel, neural network based
implementations of the GPROF algorithm:
\begin{description}
\item[GPROF-NN 1D:] This algorithm uses a fully-connected neural network to
  retrieve single column hydrometeors and rain rates based on the observed
  brightness temperature vector. Its inputs and outputs are thus identical to
  those of the standard GPROF algorithm.
\item[GPROF-NN 3D:] This algorithm extends the GPROF-NN 1D algorithm by
  incorporating spatial information into the retrieval. It produces the same
  output as GPROF and GPROF-NN 1D but processes all observations simultaneously,
  thus allowing the algorithm to combine information from pixels across the
  swath.
\end{description}
The retrieval performance of the two GPROF-NN implementations is assessed on a
held-out part of the retrieval database and compared to that of the upcoming
version of GPROF. The performance is assessed for two sensors of the GPM
constellation: The GPM Microwave Imager (GMI) and the Microwave Humidity Sounder
(MHS, \citet{bonsignori07}). In addition to this, a case study is presented that
compares the retrieved surface precipitation from overpasses of the two sensors
over Hurricane Harvey.

\section{Data and methods}

This section introduces the retrieval database that is used by the GPROF
algorithm and serves as training data for the GPROF-NN algorithms. This
is followed by a description of the current implementation of GPROF as well
as the implementation of the two GPROF-NN retrievals.

\subsection{The retrieval database}

The GPROF retrieval database consists of the retrieval inputs, i.e. PMW observations
and corresponding ancillary data, paired with corresponding retrieval outputs,
i.e. the values of the retrieval quantities. GPROF retrieves surface
precipitation, profiles and path integrals of other hydrometeors as well as
latent heating profiles. A listing of all retrieval targets and corresponding
units is provided in Tab.~\ref{tab:variables}.

\begin{table}
  \caption{Retrieval quantities in the retrieval database}
  \label{tab:variables}
  \centering
  \begin{tabular}{l|rr}
    Retrieval variable & Unit & Type\\
    \hline
    Surface precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
    Convective precipitation & $\unit{mm\ h^{-1}}$ & Scalar \\
    Cloud water path & $\unit{kg\ m^{-2}}$ & Scalar \\
    Rain water path & $\unit{kg\ m^{-2}}$ & Scalar \\
    Ice water path & $\unit{kg\ m^{-2}}$ & Scalar \\
    Cloud water content & $\unit{g\ m^{-3}}$ & Profile \\
    Rain water content & $\unit{g\ m^{-3}}$ & Profile \\
    Snow water content & $\unit{g\ m^{-3}}$ & Profile \\
    Latent heating & $\unit{K \ h^{-1}}$ & Profile \\
  \end{tabular}
\end{table}

Values of the retrieval targets for the database of the upcoming version of
GPROF are derived from one year (October 2018 to September 2019) of
hydrometeor-profile retrievals from the GPM combined product \citep{grecu16}.
This data is complemented with surface precipitation from the currently
operational MIRS retrieval by adding light precipitation in areas where no echo
is detected by the GPM Dual-Frequency Precipitation Radar. Observations over
sea-ice and snow-covered surfaces are handled separately. For sea-ice,
precipitation is derived from the ERA5 reanalysis \citep{hersbach20}. For
snow-covered surfaces, precipitation is derived from four years of co-locations
with the gauge-corrected radar observations from the Multi-Radar Multi-Sensor
(MRMS) product.

The retrieval targets are then combined with observations from each of the
sensors of the GPM constellation: For GMI, this is done by simply remapping and
averaging the retrieval outputs to the footprint of GMIs $23\ \unit{GHz}$
channel. Since co-locations with other sensors of the constellation are rare or
may not by available at all, observations for these sensors are simulated and
the footprint averaging is adapted to the expected sensor resolution.

Furthermore, ancillary data derive from global circulation models is added to
the database. It comprises two meter temperature ($T_{\text{2m}}$), total column
water vapor ($\text{TCWV}$), a surface class as well as an air lifting index
(ALI) that encodes information on atmospheric convergence in mountainous areas.
Retrieval databases are produced in two configurations for each sensor of the
GPM constellation: One for near real time applications, which uses forecasts
from the Japan Meteorological Agency (JMA) Global Analysis (GANAL) to derive
ancillary data, and one for climatological applications, which uses the ERA5
reanalysis \citep{hersbach20} to derive the ancillary data. Since this is more
of an operational detail, this study only uses databases based on ERA5.

\subsection{The GPROF algorithm}
\label{sec:gprof}

The current implementation of the GPROF retrieval consists of three components:
(1) The Bayesian method that is used to calculate the retrieval results from
retrieval inputs, (2) the compression required to handle very large databases
and (3) the incorporation of ancillary data into the retrieval. An illustration
of these three components is provided in Fig.~\ref{fig:gprof_legacy}.

\begin{figure}[hbpt!]
  \centering \includegraphics[width=\textwidth]{../figures/gprof_legacy}
  \caption{The GPROF retrieval algorithm. Panel (a) illustrates the resampling
    of database samples based on the Bayesian scheme that is the foundation of
    the GPROF retrieval. The colored, filled curve on the right side of the plot
    depicts the approximate PDF derived from the sample weights. Panel (b) shows
    the clustering of observations that is performed to speed up the calculation
    of retrieval results. Panel (c) illustrates the incorporation of ancillary
    data through the application of a stratified retrieval database.}
  \label{fig:gprof_legacy}
\end{figure}

The Bayesian scheme that is used in GPROF calculates the retrieval result as the
posterior distribution $p(\mathbf{x} | \mathbf{y})$ of the observed atmospheric
state $\mathbf{x}$ conditioned on the retrieval input $\mathbf{y}$, here a
vector of PMW observations. The database that is used for that consists of pairs
$(\mathbf{y}_1, \mathbf{x}_1), \ldots (\mathbf{y}_N, \mathbf{x}_N)$ of
observations $\mathbf{y}_i$ and states $\mathbf{x}_i$. Assuming the database is
distributed according to the the true a priori $p(\mathbf{x})$, i.e. the
climatological distribution of the atmospheric state $\mathbf{x}$, the expected
value of $\mathbf{x}$ w.r.t to the posterior distribution $p(\mathbf{x} |
\mathbf{x})$ can be approximated using
\begin{align}\label{eq:gprof_retrieval}
  \int_{\mathbf{x} } \mathbf{x} p(\mathbf{x} | \mathbf{y})\: d\mathbf{x} =
  \int_{\mathbf{x} } \mathbf{x}\: \frac{p(\mathbf{y} |
    \mathbf{x})p(\mathbf{x})}{p(\mathbf{y})}\: d\mathbf{x} \approx \frac{\sum_i
    \mathbf{x} P(\mathbf{y}|\mathbf{x}_i)}{\sum_i P(\mathbf{y}|\mathbf{x})}
\end{align}
The conditional probability $P(\mathbf{y} | \mathbf{x}_i)$ of the observation
$\mathbf{y}$ given an atmospheric state from the database is modeled assuming
unbiased Gaussian errors in the observations 
  %
\begin{align}\label{eq:gprof_error}
  P(\mathbf{y}|\mathbf{x}_i) = \frac{1}{\sqrt{2\pi\text{det}(\mathbf{S}^{-1})}}
  \exp \left \{ - \frac{1}{2} (\mathbf{y} - \mathbf{y}_i)^T \mathbf{S}^{-1}
  (\mathbf{y} - \mathbf{y}_i) \right \}
\end{align}
using a diagonal covariance matrix $\mathbf{S}$ that descibes the errors due to
sensor noise as well as other causes of deviations of real observation
$\mathbf{y}$ from the observations in the database.
Eq.~\ref{eq:gprof_retrieval} corresponds to a resampling of the states in the
database with case-specific weights calculated using Eq.~\ref{eq:gprof_error}.
This can be extended to approximate the probability density function of the
posterior distribution or to derive probabilities of certain characteristics of
the a posteriori state such as the presence of precipitation in a given
observation.

With the current size of the retrieval database, including all samples in the
sum in Eq.~\ref{eq:gprof_retrieval} would make the retrieval prohibitively slow.
To speed up the calculation observations are grouped into self-similar clusters
and only mean observation and state as well as the number of observations is
stored in the database. The clustering is accounted for in
Eq.~\ref{eq:gprof_retrieval} by multiplying the weight
$p(\mathbf{y_i}|\mathbf{x_i})$ with the number of profiles in the cluster.

Finally, in order to constrain the retrieval further than what the PMW
observations can do alone, the sum in Eq.~\ref{eq:gprof_retrieval} is limited to
profiles from similar meteorological contexts as identified by the ancillary
data. To this end the database is stratifed with respect to $T_\text{2m},
\text{TCWV}$, surface type and air lifting index. In addition to further
constraining the retrieval, this has the benefit of further speeding up the
retrieval.

\subsection{The GPROF-NN algorithms}

The principal objective that guided the design of the GPROF-NN algorithms was to
develop a neural network retrieval that provides the exact same output as GPROF
so that it can potentially replace the current implementation in a future
update. GPROF produces several outputs that are probabilistic: A probability of
precipitation as well as the mode and terciles of the posterior distribution of
precipitation. An implementation based on standard regression neural networks
would provide no principled way to produce these probabilistic outputs due to
incompatibility of (deterministic) regression with the Bayesian formulation used
by GPROF. To overcome this limitation, the implementation of the GPROF-NN
retrievals is based on quantile regression neural networks (QRNNs). As has been
shown in \citep{pfreundschuh18}, when trained on a dataset that is distributed
according to the a priori distribution of a Bayesian retrieval, QRNNs learn to
the predict quantiles of the Bayesian posterior distribution. They thus provide
a straight forward way to reconcile neural network retrievals with the Bayesian
framework that is applied by GPROF.

This approach can be generalized to predict a sequence of quantiles, from the
cumulative distribution function (CDF) of the a posteriori distribution can be
reconstructed. Since the distribution of a scalar variable is fully described by
its CDF, the predicted CDF can be used to derive any relevant statistic of the a
posteriori distribution. For the GPROF-NN retrievals the predicted CDF is used
to derive maximum a posteriori and mean surface precipitation (the latter of
which is identical to the solution that would have been obtained with common
mean squared error regression), the terciles of the posterior distribution as
well the probability of precipitation. An illustration this principle of the
GPROF-NN retrievals is provided in Fig.~\ref{fig:gprof_nn_principle}: The
retrieval employs a neural network to predict for each input pixel a vector of
values. The elements of this vector correspond to a specific squence of
quantiles of the a posteriori distribution. These quantiles are used to
reconstruct a piece-wise linear approximation of the CDF of the a posteriori
distribution from which the retrieval outputs are derived.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=1.0\textwidth]{../plots/gprof_nn_principle}
  \caption{The basic principle of the implementation of the GPROF-NN retrievals.
    A Bayesian solution of the retrieval is obtained by predicting, for each
    input pixel, a sequence of quantiles of the a posterior distribution that is
    used to reconstruct its CDF. The predicted CDF is then used to derive the
    retrieval results.}
  \label{fig:gprof_nn_principle}
\end{figure}

\subsubsection{Training objectives}
\label{sec:objectives}

A neural network can learn to predict a quantiles $\hat{x}_\tau$ of
a given conditional distribution by training it to minimize the quantile (or pinball) loss
$\mathcal{L_\tau}$ function corresponding to the quantile
fraction $\tau$ \citep{koenker01}:
\begin{align}\label{eq:quantile_loss}
  \mathcal{L}_\tau(\hat{x}_\tau, x) &= (\tau - \mathbb{I}_{x < \hat{x}_\tau})(x
  - \hat{x}_\tau),
\end{align}
where $\hat{x}_\tau$ is the predicted quantile, $x$ is a sample of the
conditional distribution from the training data and $\mathbb{I}$ is the
indicator function taking the value 1 when $x < \hat{X}$ and 0 otherwise.

This principle can be extended to a sequence of quantiles corresponding to
quantile fractions $\tau_1, \ldots, \tau_N$ by minimizing the mean of the losses
functions corresponding to each quantile fraction:
\begin{align}\label{eq:loss_function}
  \mathcal{L}_{\tau_1,\ldots, \tau_N}(\hat{\mathbf{x}}, x) &= \frac{1}{N} \sum_i
  \mathcal{L}_{\tau_i}(\hat{x}_i, x)
\end{align}
where $\hat{x}_i$ is the $i$th component of the vector of predicted quantiles
$\hat{\mathbf{x}}$.

The GPROF-NN retrievals use the loss function Eq.~\ref{eq:loss_function} with
128 equally spaced quantiles ranging from $\tau_1 = 0.001$ to $\tau_{128} =
0.999$ for all scalar retrieval variables.

Hydrometeor profiles, however, are handled differently: To reduce the number of
network outputs, the posterior mean of hydrometeor profiles is predicted
directly using mean squared error regression. Since the output of GPROF contains
only the posterior mean of the hydrometeor concentrations, it was deemed
unnecessary to predict their full posterior distribution at each level using
quantile regression.

\subsubsection{GPROF-NN 1D}

The GPROF-NN 1D algorithm was designed to be functionally identical to the
current GPROF algorithm. Just as GPROF, it uses PMW observations from a single
pixel together with corresponding ancillary data to retrieve corresponding
precipitation and hydrometeor profiles.

\begin{figure}[hbpt]
  \centering
    \includegraphics[width=\textwidth]{../plots/gprof_nn_0d_arch}
    \caption{Illustration of the neural network architecture used in the
      GPROF-NN 1D algorithm. The network consists of a common body and
      one head for each retrieval variable. Each block in body and head
      consists of fully-connected layer, layer norm and GELU activation function.
    }
  \label{fig:gprof_nn_1d}
\end{figure}

The neural network architecture used for the GPROF-NN 1D retrieval is
illustrated in Fig.~\ref{fig:gprof_nn_1d}. A single network is trained to
predict all retrieval variables listed (c.f. Tab.~\ref{tab:variables}) using the
training objectives described Sec.~\ref{sec:objectives}. The network consists of
a shared body and a separate head for each retrieved variable. Body and heads
are built-up of blocks consisting of a fully-connected layer followed by
layer normalization \citep{ba16} and GELU \citep{hendrycks16} activation
functions. During development we have experiment with different numbers of blocks
in body ($N_b$) and each of the heads ($N_h$) but found only marginal impact on
the retrieval performance and settled for a configuration with $N_b = 6$ and
$H_h = 4$.

The GPROF-NN 1D network is trained by simultaneously minimizing the sum of the
losses of all retrieval variables. The training is performed over 60 epochs
using the Adam optimizer \citep{kingma14} with an initial learning rate of
$5\cdot10^{-4}$ and a cosine annealing learning rate schedule
\citep{loshchilov16}. Warm restarts are performed after 10, 20 and 30 epochs.

\subsubsection{GPROF-NN 3D}

The GPROF-NN 3D retrieval extends the GPROF and GPROF-NN 1D algorithms by
incorporating spatial information into the retrieval. To achieve this,
the GPROF-NN employs a convolutional neural network (CNN) that performs
the retrieval for all pixels in the swath simultaneously.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=1.0\textwidth]{../plots/gprof_nn_2d_arch_compl}
  \caption{
    The neural network architecture of the GPROF-NN 3D retrieval. Illustration
    (a) displays the structure of the Xception blocks \citep{chollet17} that
    form the building blocks of the GPROF-NN 3D model. An Xception block
    consists of two depthwise separable convolutions followed by a group
    normalization layer and a GELU activation function. Illustration (b) shows
    how the Xception blocks are used in an asymmetric encoder-decoder structure
    that forms the body of the network. Output from the body is combined with
    the ancillary data to form the inputs to the separate heads that predict the
    retrieval results for each of the retrieved variables.
  }
  \label{fig:gprof_nn_3d_arch}
\end{figure}

The network architecture for the GPROF-NN 3D algorithm, illustrated in
Fig.~\ref{fig:gprof_nn_3d_arch}, consists of an asymmetric encoder-decoder
structure followed by a separate head for each retrieved variable. The stages of
the en- and decoder are built up of what we refer to here as Xception blocks
(Fig.~\ref{fig:gprof_nn_3d_arch} (a)) because they are based on the Xception
architecture introduced in \citet{chollet17}. Each block consists of two
depthwise separable convolutions with a kernel size of 3 followed by group
normalization layers with 32 groups and GELU activation functions. The first
block in each stage of the encoder additionally contains a $3\times3$
max-pooling layer with a stride of 2 following the first $3\times3$ convolution
layer. Each downsampling block in the encoder is followed by $N = 4$ standard
Xception blocks. The stages of the decoder consist of bi-linear upsampling layer
followed by a single Xception block. The network architecture was chosen in
order to keep the memory footprint of the model low enough that a moderately
deep model could still be trained on a single GPU with $24\ \unit{GB}$ of
memory.

\begin{figure}[hbpt]
  \centering
    \includegraphics[width=\textwidth]{../plots/data_augmentation}
    \caption{
      The effect of GMIs conical viewing geometry on observed features. Panel
      (a) displays geolocated observations of the $10.6\ \unit{GHz}$ channel.
      Grey squares mark equilaterals with a side length of $200\unit{km}$
      oriented along the swath. The highlighted stripe located at the swath
      center marks the region where the values of the retrieved variables are
      available. Panel (b) shows the same observations viewed as an image on a
      uniform grid. Panel (c) shows six synthetically generated training inputs
      based on two input regions marked in Panel (b). The first row shows three
      synthetic samples that simulate the effect of viewing the input in region
      A at a different position across the GMI swath. The second row shows the
      corresponding transformations for the input in region B.
    }
  \label{fig:data_augmentation}
\end{figure}

Due the special structure of the GPROF retrieval database, it was necessary to
adapt the training scheme for the GPROF-NN 3D retrieval. For the most part, the
retrieval database contains values for the retrieved variables only at the 20
central pixels of the GMI swath (illustrated by the light stripe in
Fig~\ref{fig:data_augmentation} (a)). That means that although the network
predicts the retrieval variables at all pixels in the swath, the loss can be
calculated only at the pixels where the reference data is available.
Furthermore, the viewing geometry of PMW sensors breaks the translation symmetry
that is common for digital images and which is one of the inductive biases of
CNNs \citep{goodfellow16}. The reason for this is that the distance between and
orientation of pixels varies across the swathes of the sensors. This is
illustrated in Fig.~\ref{fig:data_augmentation} for the conical viewing geometry
of GMI. When displayed on a uniform grid, one and the same shape appears
differently depending on its position across the swath. To account for this in
the training, training samples are transformed randomly to simulate the effect
of seeing a given scene at different locations across the GMI swath. These
transformations are applied randomly when a training sample is loaded thus
ensuring that the network rarely or never sees a training scene from the same
perspective. The transformations were furthermore designed to ensure that the
relative location of the pixels at which values of retrieval variables are known
varies across the full width of the swath instead of being located always at its
center. Examples of the transformed inputs are displayed in
Fig.~\ref{fig:data_augmentation} (c).

The training of the GPROF-NN 3D algorithm for other sensors than GMI is further
complicated by the fact that the retrieval database for these sensors contains
simulated observations only at the central pixels of the GMI swath (the
highlighted pixels in Fig.~\ref{fig:data_augmentation} (a)). To obtain
two-dimensional training scenes that are sufficiently wide to train a CNN, we
make use of an intermediate CNN based model to predict the simulated brightness
temperatures across the full GMI swath. The extended simulated brightness
temperatures are then remapped from the GMI viewing geometry to the viewing
geometry of the target sensor in the same way as outlined in the paragraph
above.

\section{Results}

The first part of this section presents the the evaluation GPROF and the
GPROF-NN algorithms on a held-out test data set. The test data set consists of
the observations from the retrieval database from the first three days of every
month. It should be noted that we have deliberately limited this evaluation to
data from the retrieval database in order to isolate the effect of the retrieval
algorithm from that of the database.

We conclude this section with a case study of overpasses of Hurricane Harvey.
These results are based on real observations and thus provide an indication to
what extent the performance on the retrieval database can be expected to
generalize to real observations.

\subsection{Precipitation and water columns}

As described in Tab.~\ref{tab:variables}, the scalar variables retrieved by
GPROF are surface and convective precipitation as well as the column-integrated
concentrations of cloud droplets, rain and snow, denoted as cloud water path
(CWP), rain water path (RWP) and ice water path (IWP), respectively. Scatter
plots of the retrieval results for these 5 quantities are displayed in
Fig.~\ref{fig:results_scatter_gmi} for GMI and
Fig.~\ref{fig:results_scatter_mhs} for MHS. The frequencies in all plots have
been normalized column-wise to ensure that results for high  values
remain visible.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{../plots/results_scatter_gmi}
  \caption{
    Scatter plots of scalar retrieval targets for the three retrieval algorithms
    for GMI. The row display the results for the GPROF, GPROF-NN 1D and GPROF-NN
    3D algoriths, respectively. Columns display the results for different
    retrieval targets. Frequencies in the plots have been normalized
    column-wise, i.e. per bin of the reference value.
  }
  \label{fig:results_scatter_gmi}
\end{figure}

For surface precipitation retrieved by GMI, consistent improvements in retrieval
accuracy are observed between the GPROF and GPROF-NN 1D as well as the GPROF-NN
1D and GPROF-NN 3D algorithms. The improvements are most pronounced for very
small rain rates between $10^{-2}$ and $10^{-1}\ \unit{mm\ h^{-1}}$ but are
consistent across the full range of displayed values. For convective
precipitation, the results of GPROF deviate noticeably from the diagonal.
Slightly better results are provided by the GPROF-NN 1D algorithm, while the
GPROF-NN 3D algorithm yields the best agreement with the reference data. For the
path-integrated quantities GPROF already achieves good agreement with the
reference values. Nonetheless, small but consistent improvements are observed
for the GPROF-NN 1D and GPROF-NN 3D algorithms.

\begin{figure}[hbpt]
  \centering
  \includegraphics[width=\textwidth]{../plots/results_scatter_mhs}
  \caption{
    Same as Fig.~\ref{fig:results_scatter_gmi} but for MHS.
  }
  \label{fig:results_scatter_mhs}
\end{figure}

The results for MHS, displayed in Fig~\ref{fig:results_scatter_mhs}, paint a
similar picture. Although the overall accuracy of the retrieval is lower, the
GPROF-NN 1D retrieval consistently yields more accurate results than GPROF. Also
here the GPROF-NN 3D retrieval yields further, consistent improvements compared
to the GPROF-NN 1D retrieval.

Quantitative measures of the retrieval accuracy for all scalar retrieval targets
and the three retrieval algorithms are displayed in Tab.~\ref{tab:metrics_gmi}
for GMI and Tab.~\ref{tab:metrics_mhs} in the appendix for MHS.
The tables display bias, mean absolute error (MAE), root mean squared error
(RMSE) and the mean absolute percentage error (MAPE$_{t}$) for all test samples
events with a reference value that exceeds a quantity-specific threshold $t$.
The error metrics confirm the qualitative findings from
Fig.~\ref{fig:results_scatter_gmi} and \ref{fig:results_scatter_mhs}: The neural
network implementations outperform GPROF in terms of all considered metrics.
Moreover, the GPROF-NN 3D algorithm further improves upon the performance of
the GPROF-NN 1D algorithm. Although the errors are generally larger for MHS, the
same tendency is observed.

\begin{table}[hbpt!]

  \centering
  \caption{
    Selected error metrics and standard deviations of retrieved scalar
    variables for the three implementations of the GPROF algorithm applied to
    the GMI sensor. The standard deviations were estimated using bootstrapping
    and thus provide a measure of the uncertainty due to random sampling of the
    test data.
  }
  \label{tab:metrics_gmi}
\begin{subtable}{\textwidth} 
 \caption{Surface precipitation} 
\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
\hline
Metric &
\multicolumn{1}{|c}{GPROF} &
\multicolumn{1}{|c}{GPROF-NN 1D} &
\multicolumn{1}{|c|}{GPROF-NN 3D} \\
\hline\hline
Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0091 \pm 0.0001$ &\hfill $ -0.0019 \pm 0.0001$ &\hfill $ -0.0012 \pm 0.0001$ \\
MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0787 \pm 0.0001$ &\hfill $  0.0579 \pm 0.0000$ &\hfill $  0.0462 \pm 0.0001$ \\
RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.4049 \pm 0.0078$ &\hfill $  0.3799 \pm 0.0078$ &\hfill $  0.3398 \pm 0.0078$ \\
MAPE$_{0.1}$ \hfill [$\unit{\%}$] & \hfill $ 64.3838 \pm 0.0681$ &\hfill $ 58.0879 \pm 0.0247$ &\hfill $ 48.1921 \pm 0.0449$ \\
\hline
\end{tabular}
\end{subtable}

\begin{subtable}{\textwidth} 
 \caption{Convective precipitation} 
\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
\hline
Metric &
\multicolumn{1}{|c}{GPROF} &
\multicolumn{1}{|c}{GPROF-NN 1D} &
\multicolumn{1}{|c|}{GPROF-NN 3D} \\
\hline\hline
Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $ -0.0026 \pm 0.0001$ &\hfill $ -0.0008 \pm 0.0001$ &\hfill $ -0.0004 \pm 0.0000$ \\
MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0332 \pm 0.0002$ &\hfill $  0.0232 \pm 0.0000$ &\hfill $  0.0228 \pm 0.0001$ \\
RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.3946 \pm 0.0094$ &\hfill $  0.3550 \pm 0.0094$ &\hfill $  0.3104 \pm 0.0094$ \\
MAPE$_{0.1}$ \hfill [$\unit{\%}$] & \hfill $ 90.7868 \pm 0.1886$ &\hfill $ 88.3793 \pm 0.1458$ &\hfill $ 82.5167 \pm 0.1474$ \\
\hline
\end{tabular}
\end{subtable}

\begin{subtable}{\textwidth} 
 \caption{Rain water path} 
\begin{tabular}{|c||p{3.5cm}|p{3.5cm}|p{3.5cm}|}
\hline
Metric &
\multicolumn{1}{|c}{GPROF} &
\multicolumn{1}{|c}{GPROF-NN 1D} &
\multicolumn{1}{|c|}{GPROF-NN 3D} \\
\hline\hline
Bias \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0009 \pm 0.0000$ &\hfill $ -0.0004 \pm 0.0000$ &\hfill $ -0.0002 \pm 0.0000$ \\
MAE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0167 \pm 0.0000$ &\hfill $  0.0118 \pm 0.0000$ &\hfill $  0.0089 \pm 0.0000$ \\
RMSE \hfill [$\unit{mm\ h^{-1}}$] & \hfill $  0.0913 \pm 0.0032$ &\hfill $  0.0955 \pm 0.0032$ &\hfill $  0.0702 \pm 0.0032$ \\
MAPE$_{0.01}$ \hfill [$\unit{\%}$] & \hfill $ 78.5511 \pm 0.1381$ &\hfill $ 69.4666 \pm 0.0849$ &\hfill $ 59.6460 \pm 0.0470$ \\
\hline
\end{tabular}
\end{subtable}
